The resurgence of the linear optics interferometer --- recent advances \& applications

### 2017-09-10 paper.tex
Added a section on quantum simulation

\section{Reconstructing the linear optical network}\label{sec:reconstruction}

In many practical situations, the structure of a linear optical device in terms of its constituent beamsplitters and phase-shifters is known once it is built. However, owing to manufacturing imperfections, a precise characterization of these devices may still be needed post-production. One approach for achieving this is via quantum process tomography.
%\cite{bib:Mitchell03,bib:Obrien04,bib:Lobino08,bib:Saleh11}.
However, quantum process tomography is an expensive approach in terms of the number of measurements required to characterize the network, with exponential overhead, becoming impractical for large optical networks which can contain hundreds of modes using present-day technology \cite{bib:Harris16}. To mitigate this problem, alternative characterization protocols have been developed.

Generally, the unitary matrices of $d\times d$ linear optical devices are complex $U_{ij}=r_{ij}e^{i\theta_{ij}}$, where $0 \leq r_{ij}\leq 1$, and $0\leq \theta_{ij}\leq 2\pi$. To characterize these numbers, one can do so by injecting one- and two-photon states into the network with correlated photon detection \cite{bib:Laing12}. With some mathematical simplifications, the parameters can be solved for using the one-photon transmissions, and the visibility of two-photon inputs.
%
%First, they note some equivalencies: two unitaries $U$ and $U'$ are 
%equivalent if there exist two diagonal unitary matrices $D_1^U$ and 
%$D_2^U$ such that $U'=D^U_1 U D^U_2$, because these diagonal matrices are 
%regarded as unknown and trivial phases on the input and output ports of 
%the network, to which the one- and two-photon data are insensitive to. 
%This reduces the first row and column elements to real numbers, i.e. $
%\theta_{1j}=\theta_{i1}=0$. Second, the photo-statistics remain invariant 
%under complex conjugation by $U$. Thus, the imaginary part of $M_{2,2}$ 
%must be non-negative. Then, assuming the first two rows and columns are 
%non-vanishing, and there is no total loss in the interferometer, the 
%matrix to be recovered is
%\begin{align}
%U=\left (\begin{array}{cccc}
%r_{11} & r_{12} & \ldots & r_{1m} \\
%r_{21} & r_{22}e^{i\theta_{22}} &\ldots & r_{2m}e^{i\theta_{2m}}\\
%\vdots & \vdots & \ddots & \vdots \\
%r_{m1} & r_{m2}e^{i\theta_{m2}} & \ldots & r_{mm}e^{i\theta_{mm}}
% \end{array}\right ) \ .
%\end{align}
%They showed it is possible to write the parameters  for $i,j\geq 2$  The
% remaining $2m-1$ real parameters can be found via one-photon 
%transmissions.
%
An increased accuracy in the characterization is possible by estimating 
and correcting systematic errors that arise due to mode mismatch 
\cite{Dhand16}. Others have used numerical methods to find the closest 
parameters that yield the observed visibilities 
\cite{bib:Spagnolo16,bib:Tillmann16}.

An alternative method uses coherent states to probe the interferometer \cite{bib:Rahimi-Keshari13,bib:Heilmann15} instead of Fock states. Such states are produced by a standard laser source, thus reducing experimental resources. The $r_{jk}$ terms can be calculated from the ratios of output intensities at the $k$th port to the input intensity at the $j$th port. The remaining phases $\theta_{ij}$ are found by the interference pattern given by a two-mode coherent state $\ket{\alpha}\ket{\alpha e^{\phi}}$. When the states are injected into ports 1, and $j$ respectively, the output intensity at the $k$th port is
\begin{align}
I_k=I(r_{1k}^2+r_{jk}^2+2 r_{1k}r_{kj}\cos(\phi+\theta_{jk})) \ ,
\end{align}
where $\theta_{jk}=0$ for $k=1$, and $I$ is the intensity of the input coherent states. By scanning the phase shift $\phi$ and locating the maximum value of $I_k$ for $j=2,\ldots, m$, all unknown phases can be found via $\theta_{jk}=2\pi-\phi$. An elegant modification of this scheme removes the need for precise control of the phase-shift $\phi$ by suggesting instead to plot the output intensity $I_k$ with respect to the input intensity $I$ \cite{bib:Heilmann15}. In time, the natural drift in the laser source will cause this plot to trace out an ellipse, known as a Lissajous figure, whose orientation and direction of evolution will give the phase $\theta_{jk}$ and its sign respectively.

@@

The difficulty in producing nonlinear operations required for universal LOQC led researchers to seek out alternative ways for demonstrating its power in the near term. {\sc Bosonsampling} is the result of such an effort. It is a restricted model of non-universal quantum computation \cite{bib:Aaronson11} that constitutes sampling the photocounts from the probability distribution of identical bosons scattered by a passive linear optical interferometer. While not universal, the {\sc bosonsampling} scheme is strongly believed to implement a classically hard task. Consider such a circuit of $m$ input (and output) modes that is injected with $n$ indistinguishable single photons such that $m\sim \mathcal{O}(n^2)$. Then, the boson-sampling task consists of generating a sample from the probability distribution of single-photon measurements at the output of the circuit. The probability of detecting $s_j$ photons at the $j$th output mode is
\begin{align}
p(s_1,\ldots,s_m)=\frac{|{\rm Per}(U_S)|^2}{s_1!\ldots s_n!} \ ,	
\end{align}
where ${\rm Per}()$ denotes the permanent of a matrix, $S=\{s_1,\dots,s_m\}$ denotes an output photon-number configuration, and $U_S$ is obtained from the interferometer unitary $U$ by keeping its first $n$ columns and repeating its $j$th row $s_j$ times. The appearance of the permanent, which is \#P-hard in general (a complexity class even harder than NP-hard), in this statistics contributes to the hardness of the boson-sampling problem. Note that the number of configurations $|S|=\binom{n+m-1}{n}$ in the output superposition scales exponentially with $m$. For this reason, the boson sampler does not let us \emph{calculate} matrix permanents, as this would require knowing individual amplitudes with high precision, which would require an exponential number of samples. Rather, the boson sampler samples across a distribution of permanents without actually revealing any of them to the experimenter. Nonetheless, this is a classically intractable problem. It is also known that boson-sampling is as hard as stimulating the short time evolution of a certain Hamiltonian\cite{bib:Peropadre17}.

@@

@@

 As discussed in Section \ref{sec:bosonsampling}, the resources needed for Boson-sampling is readily available, but executes a task that is believed to be hard classically. Surprisingly, an entanglement between a large number of path labels is possible with single photons in a passive linear optical device. Motes {\it et al.~}harnessed this so-called number-path entanglement for quantum metrology without the use of entangled states \cite{bib:Motes15_LOQM}. Here, the phase shift to be estimated, $\varphi$, manifests itself as a linear phase gradient across $N$ modes (See Fig. \ref{fig:MORDOR}). By straddling these modes among a quantum Fourier transform and its conjugate operation, it is possible to achieve $\Delta \varphi=\mathcal{O}(N^{-3/2})$ using a coincidence measurement with just a string of $N$ single photons, for up to at least $N=25$ modes.

@@

>>>>

\section{Reconstructing linear optics networks}\label{sec:reconstruction}

\peter{
In many practical situations, the structure of a linear optical device in terms of its constituent beamsplitters and phase-shifters is known once it is built. However, owing to manufacturing imperfections, a precise characterization of these devices may still be needed post-production. One approach for achieving this is via quantum process tomography, discussed in the next section.

However, quantum process tomography is an expensive approach in terms of the number of measurements required to characterise the network, with exponential overhead, becoming impractical for large optical networks which can contain hundreds of modes using present-day technology \cite{bib:Harris16}. To mitigate this problem, alternative characterisation protocols have been developed.

Generally, the unitary matrices of $d\times d$ linear optical devices are complex $U_{ij}=r_{ij}e^{i\theta_{ij}}$, where $0 \leq r_{ij}\leq 1$, and $0\leq \theta_{ij}\leq 2\pi$. To characterise these numbers, one can do so by injecting one- and two-photon states into the network with correlated photon detection \cite{bib:Laing12}. With some mathematical simplifications, the parameters can be solved for using the one-photon transmissions, and the visibility of two-photon inputs.
%
%First, they note some equivalencies: two unitaries $U$ and $U'$ are 
%equivalent if there exist two diagonal unitary matrices $D_1^U$ and 
%$D_2^U$ such that $U'=D^U_1 U D^U_2$, because these diagonal matrices are 
%regarded as unknown and trivial phases on the input and output ports of 
%the network, to which the one- and two-photon data are insensitive to. 
%This reduces the first row and column elements to real numbers, i.e. $
%\theta_{1j}=\theta_{i1}=0$. Second, the photo-statistics remain invariant 
%under complex conjugation by $U$. Thus, the imaginary part of $M_{2,2}$ 
%must be non-negative. Then, assuming the first two rows and columns are 
%non-vanishing, and there is no total loss in the interferometer, the 
%matrix to be recovered is
%\begin{align}
%U=\left (\begin{array}{cccc}
%r_{11} & r_{12} & \ldots & r_{1m} \\
%r_{21} & r_{22}e^{i\theta_{22}} &\ldots & r_{2m}e^{i\theta_{2m}}\\
%\vdots & \vdots & \ddots & \vdots \\
%r_{m1} & r_{m2}e^{i\theta_{m2}} & \ldots & r_{mm}e^{i\theta_{mm}}
% \end{array}\right ) \ .
%\end{align}
%They showed it is possible to write the parameters  for $i,j\geq 2$  The
% remaining $2m-1$ real parameters can be found via one-photon 
%transmissions.
%
An increased accuracy in the characterization is possible by estimating 
and correcting systematic errors that arise due to mode mismatch 
\cite{Dhand16}. Others have used numerical methods to find the closest 
parameters that yield the observed visibilities 
\cite{bib:Spagnolo16,bib:Tillmann16}.

An alternative method to process tomography is feeding coherent states to probe the interferometer \cite{bib:Rahimi-Keshari13,bib:Heilmann15} instead of Fock states. Such states are produced by a standard laser source, thus reducing experimental resources. The $r_{jk}$ terms can be calculated from the ratios of output intensities at the $k$th port to the input intensity at the $j$th port. The remaining phases $\theta_{ij}$ are found by the interference pattern given by a two-mode coherent state $\ket{\alpha}\ket{\alpha e^{\phi}}$. When the states are injected into ports 1, and $j$ respectively, the output intensity at the $k$th port is
\begin{align}
I_k=I(r_{1k}^2+r_{jk}^2+2 r_{1k}r_{kj}\cos(\phi+\theta_{jk})) \ ,
\end{align}
where $\theta_{jk}=0$ for $k=1$, and $I$ is the intensity of the input coherent states. By scanning the phase shift $\phi$ and locating the maximum value of $I_k$ for $j=2,\ldots, m$, all unknown phases can be found via $\theta_{jk}=2\pi-\phi$. An elegant modification of this scheme removes the need for precise control of the phase-shift $\phi$ by suggesting instead to plot the output intensity $I_k$ with respect to the input intensity $I$ \cite{bib:Heilmann15}. In time, the natural drift in the laser source will cause this plot to trace out an ellipse, known as a Lissajous figure, whose orientation and direction of evolution will give the phase $\theta_{jk}$ and its sign respectively.

While the input-to-output amplitudes of a linear optics network may be efficiently determined (in terms of number of measurements), it does not characterise the full quantum process it implements, which may contain errors in other degrees of freedom.
}

\section{Quantum state and process tomography}\label{secQPT}

\peter{When demonstrating quantum protocols experimentally it is generally absolutely critical to characterise the device prior to actual protocol execution. Tomography can be used to reconstruct the full density operator of an output state, and quantum process tomography to reconstruct the full process matrix implemented by a device. While various other measurement schemes can partially reconstruct quantum states or processes, or simply reconstruct certain parameters quantifying them, tomography is complete -- fully and uniquely reconstructing state and process matrices, from which any other measure can be determined. Importantly, state and process tomography require no interference or multi-qubit gates -- just separable single-qubit measurements in different bases. But unfortunately they require a number of measurements growing exponentially with the size of the system.

State tomography is achieved by measuring expectation values in a complete basis for the state Hilbert space. For one qubit, for example, a complete basis is the Pauli basis, which wields the completeness relation,}
\begin{align}
\hat\rho = \text{tr}({\hat\rho}) + \text{tr}(\hat{X}\hat{\rho})\hat{X} + \text{tr}({\hat{Y}\hat\rho})\hat{Y} + \text{tr}(\hat{Z}{\hat\rho})\hat{Z}.
\end{align}
\peter{These traces correspond exactly to the expectation values in the respective bases, which are easily determined using separable measurements. Process tomography employs state tomography as a primitive, applying an equivalent completeness relation, but across the space of quantum processes, which may be decomposed into a basis input to output states. For a single qubit process this may be represented as,}
\begin{align} \label{eq:process_matrix}
\mathcal{E}(\hat\rho) = \sum_{i,j=1}^4 \chi_{i,j} \hat{\sigma}_i\hat\rho\,\hat{\sigma}_j^\dag.
\end{align}
\peter{which generalises to an arbitrary number of qubits.Here $\chi$ is the process matrix in the Pauli basis.}
%\cite{bib:Mitchell03,bib:Obrien04,bib:Lobino08,bib:Saleh11}.

@@

The difficulty in producing nonlinear operations required for universal LOQC led researchers to seek out alternative ways for demonstrating its power in the near term. {\sc Bosonsampling} is the result of such an effort. It is a restricted model of non-universal quantum computation \cite{bib:Aaronson11} that constitutes sampling the photocounts from the probability distribution of identical bosons scattered by a passive linear optical interferometer. While not universal, the {\sc bosonsampling} scheme is strongly believed to implement a classically hard task. Consider such a circuit of $m$ input (and output) modes that is injected with $n$ indistinguishable single photons such that $m\sim \mathcal{O}(n^2)$. Then, the boson-sampling task consists of generating a sample from the probability distribution of single-photon measurements at the output of the circuit. The probability of detecting $s_j$ photons at the $j$th output mode is
\begin{align}
p(s_1,\ldots,s_m)=\frac{|{\rm Per}(U_S)|^2}{s_1!\ldots s_n!} \ ,	
\end{align}
%<<<<<<< HEAD:paper.tex
where $\text{Per}()$ denotes the permanent of a matrix, $S=\{s_1,\dots,s_m\}$ denotes an output photon-number configuration, and $U_S$ is obtained from the interferometer unitary $U$ by keeping its first $n$ columns and repeating its $j$th row $s_j$ times. The appearance of the permanent, which is \#P-hard in general (a complexity class even harder than NP-hard), in this statistics contributes to the hardness of the boson-sampling problem. Note that the number of configurations $|S|=\binom{n+m-1}{n}$ in the output superposition scales exponentially with $m$. For this reason, the boson sampler does not let us \emph{calculate} matrix permanents, as this would require knowing individual amplitudes with high precision, which would require an exponential number of samples. Rather, the boson sampler samples across a distribution of permanents without actually revealing any of them to the experimenter. Nonetheless, this is a classically intractable problem. It is also known that boson-sampling is as hard as stimulating the short time evolution of a certain Hamiltonian\cite{bib:Peropadre17}.

where ${\textbf Per}()$ denotes the permanent of a matrix, $S=\{s_1,\dots,s_m\}$ denotes an output photon-number configuration, and $U_S$ is obtained from the interferometer unitary $U$ by keeping its first n columns and repeating its $j$th row $s_j$ times. The appearance of the permanent, which is \#P-hard in general (a complexity class even harder than NP-hard), in this statistics contributes to the hardness of the boson sampling problem. Note that the number of configurations $|S|=\binom{n+m-1}{n}$ in the output superposition scales exponentially with $m$. For this reason, the boson sampler does not let us \emph{calculate} matrix permanents, as this would require knowing individual amplitudes with high precision, which would require an exponential number of samples. Rather, the boson sampler samples across a distribution of permanents without actually revealing any of them to the experimenter. Nonetheless, this is a classically intractable problem. It is also known that boson sampling is as hard as stimulating the short time evolution of a certain Hamiltonian\cite{bib:Peropadre17}.
%>>>>>>> origin/master:(PR final edits) linear_optics_review/paper.tex

@@

\subsection{Other sampling problems -- hard \& easy}

\peter{
Numerous other sampling problems have since been described, using, for example,
\begin{enumerate}
	\item Cat states (superpositions of coherent states)
	\item Photon-added coherent states .
	\item Photon-added or -subtracted squeezed vacuum states.
	\item Two-mode squeezed vacuum states (SPDC states).
	\item Gaussian states.
	\item Positive P-function states.
\end{enumerate}
}

\peter{
Of these, 1-4 have been shown to yield hard sampling problems under photodetection, assuming certain reasonable complexity assumptions. And 5-6 have been shown to implement polynomial-time classical algorithms, under Gaussian measurements.}

\sihui{
\subsection{Quantum simulations}
Quantum simulations is about simulating a quantum system by quantum mechanical means \cite{bib:Georgescu14}, and there are generally two approaches to achieving this. The first approach is known as digital quantum simulation and achieved by building a quantum computer from unitary gate components that are generated by a Hamiltonians of quantum processes. Progress using this approach has already been discussed in Section \ref{sec:LOQC}.The second approach is known as analog quantum simulation and involves building simpler analog devices that mimic other, perhaps less accessible, quantum systems. We focus on discussing analog quantum simulation in this section.

 Quantum interferometers have been proposed and in some cases implemented as quantum simulators. In quantum chemistry, the energy spectrum of the hydrogen molecule has been calculated using a quantum simulator comprising of a pair of entangled photons in a polarization-dependent interferometer \cite{bib:Lanyon10}. With two entangled photon pairs, valence-bond states have been simulated and a fiber-based tunable directional coupler gives analogue control over superpositions of different valence-bond states \cite{bib:Ma2011}. The model of {\sc Bosonsampling}, when coupled with two-level systems that are coupled to the ports of the quantum interferometer, simulates a general spin XY Hamiltonian \cite{bib:Olivares16}. Quantum interferometers can also be used to simulate quantum walks which in turn can be used to simulate tight-binding Hamiltonians,  bosonic-to-fermionic behavior transitions, and even anyonic behavior.
  }

@@

 As discussed in Section \ref{sec:bosonsampling}, the resources needed for Boson-sampling \peter{are} readily available, but executes a task that is believed to be hard classically. Surprisingly, an entanglement between a large number of path labels is possible with single photons in a passive linear optical device. Motes {\it et al.~}harnessed this so-called number-path entanglement for quantum metrology without the use of entangled states \cite{bib:Motes15_LOQM}. Here, the phase shift to be estimated, $\varphi$, manifests itself as a linear phase gradient across $N$ modes (See Fig. \ref{fig:MORDOR}). By straddling these modes among a quantum Fourier transform and its conjugate operation, it is possible to achieve $\Delta \varphi=\mathcal{O}(N^{-3/2})$ using a coincidence measurement with just a string of $N$ single photons, for up to at least $N=25$ modes.

@@

\section{Optical switching}

\peter{
Some of the protocols thus far require switching between different optical modes. This is achieved experimentally via a Mach-Zehnder interferometer with a dynamic, classically-controlled phase-shift in one path. The variable phase may be implemented in a number of ways using various types of phase modulators, such as electro-optic or acousto-optic modulators. With this construction setting the phase to $\phi=0$ performs the identity operation, whereas for $\phi=\pi$ it performs a swap gate. This is shown in Fig. \ref{figMZ}.
}

\begin{figure}[!htb]
\centering
\includegraphics[width=0.55\columnwidth]{MZ}	
\caption{} \label{figMZ}
\end{figure}

\peter{
This concept for switching is widely used in current labs, however it is very slow using present-day technology. For the purposes of fast-feedforward -- required for universal optical quantum computing for example -- we would require switching times on the order of nanoseconds, which requires the switching time to be faster than the time to the destination of the light it needs to switch. Presently this remains a daunting challenge, requiring much further research, with quantum memories being a candidate for loosening the switching time limitations.
}
